Learning Reusable Components for Neurally--Guided Bayesian Program Learning

### 2018-05-13 nips_2018.tex
shortened: we might actually make it within the page limit...

\system takes as input a set of \emph{tasks}, written $X$, each of which is a program induction problem.
It has at its disposal a \emph{likelihood model}, written $\probability[x|p]$, which scores the likelihood of a task $x\in X$ given a program $p$.
Its goal is to solve each of the tasks by writing a program,
and also to infer a DSL, written $\mathcal{D}$.
We equip $\mathcal{D}$ with a real-valued weight vector $\theta$, and together
$(\mathcal{D},\theta)$ define a generative model over programs.
We frame our goal as maximum a posteriori (MAP) inference of $(\mathcal{D},\theta)$ given $X$.
Writing $J$ for the joint probability of $(\mathcal{D},\theta)$ and $X$, we want the $\mathcal{D}^*$ and $\theta^*$ solving:
\begin{align}\label{intractableObjectives}
  J(\mathcal{D},\theta)\triangleq \probability[\mathcal{D},\theta]\prod_{x\in X} \sum_p \probability[x|p]\probability[p|\mathcal{D},\theta],\quad
  \mathcal{D}^* = \argmax_{\mathcal{D}}\int J(\mathcal{D},\theta)\;\mathrm{d}\theta, \quad
  \theta^* =\argmax_\theta J(\mathcal{D}^*,\theta)
\end{align}

@@

Alg.~\ref{programGenerativeModel} is a procedure for drawing
samples from the generative model $(\mathcal{D},\theta)$.  In practice, we
enumerate programs in order of their probability under  Alg.~\ref{programGenerativeModel} rather than sampling them.
\begin{wrapfigure}{L}{0.5\textwidth}
  \begin{minipage}{0.49\textwidth}    
\begin{algorithm}[H]
   \caption{Generative model over programs}
   \label{programGenerativeModel}
   \begin{algorithmic}
%%      \STATE \textbf{function} sampleProgramFromDSL$(\mathcal{D}, \theta, \tau)$:
%%   \STATE {\bfseries Input:} DSL $\mathcal{D}$, weight vector $\theta$, type $\tau$
%%   \STATE \textbf{Output:} a program whose type unifies with $\tau$
%%   \STATE \textbf{return} sample$(\mathcal{D}, \theta, \varnothing, \tau)$
%% \STATE
     \STATE \textbf{function} sample$(\mathcal{D}, \theta, \mathcal{E}, \tau)$:
  \STATE {\bfseries Input:} DSL $(\mathcal{D},\theta)$, environment $\mathcal{E}$, type $\tau$
  \STATE \textbf{Output:} a program whose type unifies with $\tau$
  \IF{$\tau = \alpha\to\beta$}
  \STATE var $\gets$ an unused variable name
  \STATE body $\sim$ sample$(\mathcal{D},\theta,\{\text{var}:\alpha\}\cup\mathcal{E},\beta)$
   \STATE \textbf{return} \code{(lambda (}var\code{) }body\code{)}
   \ENDIF
   %   \ELSE
   \STATE $\text{primitives} \gets\{p | p: \tau' \in \mathcal{D}\cup\mathcal{E}$ \\
      \hspace*{6.9em}$\text{if }\tau\text{ can unify with yield}(\tau') \} $

   \STATE Draw $e\sim \text{primitives}$, w.p. $\propto\theta_e$ if $e\in \mathcal{D}$ \\
      \hspace*{8.8em}w.p. $\propto\frac{\theta_{var}}{|\text{variables}|}$ if $e\in \mathcal{E}$
   \STATE Unify $\tau$ with yield$(\tau')$.
   \STATE $\left\{\alpha_k \right\}_{k = 1}^K\gets\text{args}(\tau')$ 
%   \STATE unify$(\tau,\beta)$
   \FOR{$k=1$ {\bfseries to} $K$}
 \STATE $a_k\sim\text{sample}(\mathcal{D},\theta,\mathcal{E},\alpha_k)$
 \ENDFOR
 \STATE \textbf{return} \code{(}$e\;a_1\; a_2\; \cdots\; a_K$\code{)}
 \STATE\textbf{where:}
 \STATE yield$(\tau) = \begin{cases}
   \text{yield}(\beta)   &\text{ if }\tau = \alpha\to \beta\\
   \tau   &\text{ otherwise.}
 \end{cases}$ 
 \STATE  args$(\tau) = \begin{cases}
   [\alpha] + \text{args}(\beta)   &\text{ if }\tau = \alpha\to \beta\\
   []   &\text{ otherwise.}
 \end{cases}$
\end{algorithmic}
\end{algorithm}
    \end{minipage}
  \end{wrapfigure}

@@

We want a recognition model that closely approximates the true posteriors over programs. We formulate this as minimizing the expected KL-divergence, $  \expect\left[\text{KL}\left(\probability[p|x,\mathcal{D},\theta]\|q(p|x) \right) \right]$,
or equivalently maximizing $  \expect\left[\sum_p\probability[p|x,\mathcal{D},\theta]\log q(p|x) \right]$,
%% \begin{equation*}
%%   \expect\left[\sum_p\probability[p|x,\mathcal{D},\theta]\log q(p|x) \right]
%% \end{equation*}
where the expectation is taken over tasks. One could take this expectation
over the observed empirical distribution of tasks,
like how an autoencoder is trained~\cite{hinton2006reducing}; or, one could take this expectation over samples from the generative model, like how a Helmholtz machine is trained~\cite{dayan1995helmholtz}.
We found it useful to maximize both an autoencoder-style objective $\mathcal{L}_{\text{AE}}$ and a Helmholtz-style objective $\mathcal{L}_{\text{HM}}$, giving the  objective for a recognition model, $\mathcal{L}_{\text{RM}} = \mathcal{L}_\text{AE} + \mathcal{L}_\text{HM}$:
\begin{align}\nonumber
\mathcal{L}_{\text{HM}} = \expect_{p\sim(\mathcal{D},\theta) }\left[\log q(p|x)\right]\text{ s.t. $x = \sem{p}$}\quad
\mathcal{L}_{\text{AE}} = \expect_{x\sim X}\left[\sum_{p\in \mathcal{F}_x}
  \frac{\probability\left[x,p|\mathcal{D},\theta \right]}{\sum_{p'\in \mathcal{F}_x}\probability\left[x,p'|\mathcal{D},\theta \right]}\log q(p|x)\right]
\end{align}
The $\mathcal{L}_{\text{HM}}$ objective is essential for data efficiency:
all of our experiments train \system on only a few hundred tasks, which is too little for
a high-capacity neural network $q$.
Once we bootstrap a $(\mathcal{D},\theta)$,
we can draw unlimited samples from $(\mathcal{D},\theta)$
and train $q$ on those samples.

@@

We define a prior distribution over DSLs which penalizes the sizes of the $\lambda$-calculus expressions in the DSL, and put a Dirichlet prior over the weight vector:
\begin{equation}
    \probability[\mathcal{D}]\propto\exp\left(-\lambda\sum_{p\in \mathcal{D}}\text{size}(p) \right)\hspace{1cm}  \probability[\theta|\mathcal{D}] = \text{Dir}(\theta|\alpha)
  \end{equation}

where $\text{size}(p)$  measures the size of the syntax tree of program $p$,
$\lambda$ is a hyperparameter that acts as a regularizer on the size of the DSL,
and $\alpha$ is a concentration parameter controlling the smoothness of the prior over $\theta$.
%Alg.~\ref{grammarInductionAlgorithm} specifies the DSL induction algorithm.

To appropriately score each proposed $\mathcal{D}$ we must reestimate the weight vector $\theta$.
Although this problem may seem 
very similar to estimating the parameters of a probabilistic context free grammar (PCFG),
for which we have effective approaches like the Inside/Outside algorithm~\cite{international2000derivation},
a DSL in \system may be context-sensitive due to the presence of variables
in the programs and also due to the polymorphic typing system.
In the Supplement we derive a tractable MAP estimator for $\theta$.

\begin{figure*}
\centering  \begin{tabular}{ll}
    \toprule
    Example programs in frontiers&Proposed subexpression\\\midrule
    \begin{tabular}{l}
      \code{(lambda (a b) (fold b (cons "," a)}\\
      \phantom{\code{(lambda }}\code{(lambda (x z) (cons x z))))}\\
      \code{(lambda (a b) (fold a b (lambda (x z) (cons x z))))}
    \end{tabular}&
    \begin{tabular}{l}
      \code{(fold a b (lambda (x z) }\\
      \phantom{\code{(fold a b }}\code{(cons x z)))}
      \end{tabular}
\\\bottomrule
  \end{tabular}
  \caption{The DSL induction algorithm proposes subexpressions of programs to add to the DSL.
    These subexpressions are taken from programs in the frontiers (left column), and can introduce new variables (right column: \code{a} and \code{b}). Here, the proposed subexpression appends two lists.} 
  \label{fragmentExample}
\end{figure*}

Alg.~\ref{mainAlgorithm} describes how we combine program search,
recognition model training, and DSL induction.
\begin{comment}
\subsection{Implementing \system}

Alg.~\ref{mainAlgorithm} describes how we combine program search,
recognition model training, and DSL induction.
We note the following implementation details:
(1) We perform an exploration cycle before each compression and compilation  cycle.
(2) On the first iteration, we do \emph{not} train the recognition model on samples from
the generative model because a generative model has not yet been learned --
we instead train the network to only maximize $\mathcal{L}_{\text{AE}}$.
%% (3) Because the frontiers can grow very large,
%% we only keep around the top $10^4$ programs $p$ in each frontier $\mathcal{F}_x$
%% with the highest likelihood $\probability[x,p|\mathcal{D},\theta]$.
(3) During both DSL induction and neural net training, we calculate Eq.~\ref{AIC} and $\mathcal{L}_{\text{RM}}$
by only summing over the top $K$
programs in $\mathcal{F}_x$ as measured by $\probability[x,p|\mathcal{D},\theta]$ -- we found that $K = 2$ sufficed.
(4) For added robustness, we enumerate programs from both the generative model $(\mathcal{D},\theta)$ and the recognition model.
\end{comment}
\begin{wrapfigure}{L}{0.6\textwidth}
  \begin{minipage}{0.6\textwidth}    
\begin{algorithm}[H]
   \caption{The \system Algorithm}
   \label{mainAlgorithm}
   \begin{algorithmic}
     \STATE {\bfseries Input:} Initial DSL $\mathcal{D}$, set of tasks $X$, iterations $I$
     \STATE \textbf{Hyperparameters:} Enumeration timeout $T$
     \STATE \textbf{Output:} DSL $\mathcal{D}$, weight vector $\theta$, recognition model $q(\cdot)$
     \STATE Initialize $\theta\gets \text{uniform}$ %, $q_0(\cdot ) = \theta_0$
     \FOR{$i=1$ {\bfseries to} $I$}
     %     \FOR{$x:\tau\in X$}
     \STATE  $\mathcal{F}^{\theta}_x\gets \{p| p\in \text{enum}(\mathcal{D},\theta,T)\text{ if }\probability[x|p] > 0\}$ \footnotesize{(\textbf{Explore})}
     \STATE $q\gets \text{train recognition model, maximizing }\mathcal{L}_{\text{RM}}$ \hspace{0.2cm}(\footnotesize{\textbf{Compile}})
     \STATE  $\mathcal{F}^{q}_x\gets\{p|p\in \text{enum}(\mathcal{D},q(x),T)\text{ if }\probability[x|p] > 0\}$ \footnotesize{\textbf{(Explore)}}

     \STATE $\mathcal{D},\theta\gets $induceDSL$(\{\mathcal{F}^{\theta}_x\cup\mathcal{F}^{q}_x\}_{x\in X})$  \hspace{1.1cm} \footnotesize{\textbf{(Compress)}}
     %% \STATE Define $Q_x(z) \propto \begin{cases}
     %%   \probability[x|z]\probability[z|\mathcal{D}_i,\theta_i]&x\in \mathcal{F}_x\\
     %%   0&x\not \in \mathcal{F}_x
     %% \end{cases}
      \ENDFOR
 \STATE \textbf{return} $\mathcal{D},\theta,q$
\end{algorithmic}
\end{algorithm}
    \end{minipage}
  \end{wrapfigure}

>>>>

\system takes as input a set of \emph{tasks}, written $X$, each of which is a program induction problem.
It has at its disposal a \emph{likelihood model}, written $\probability[x|p]$, which scores the likelihood of a task $x\in X$ given a program $p$.
Its goal is to solve each of the tasks by writing a program,
and also to infer a DSL, written $\mathcal{D}$.
We equip $\mathcal{D}$ with a real-valued weight vector $\theta$, and together
$(\mathcal{D},\theta)$ define a generative model over programs.
We frame our goal as maximum a posteriori (MAP) inference of $(\mathcal{D},\theta)$ given $X$.
Writing $J$ for the joint probability of $(\mathcal{D},\theta)$ and $X$, we want the $\mathcal{D}^*$ and $\theta^*$ solving:
\begin{align}\label{intractableObjectives}
\nonumber  J(\mathcal{D},\theta)\triangleq \probability[\mathcal{D},\theta]\prod_{x\in X} \sum_p \probability[x|p]\probability[p|\mathcal{D},\theta]\\
  \mathcal{D}^* = \argmax_{\mathcal{D}}\int J(\mathcal{D},\theta)\;\mathrm{d}\theta \qquad
  \theta^* =\argmax_\theta J(\mathcal{D}^*,\theta)
\end{align}

@@

\begin{wrapfigure}{L}{0.5\textwidth}
  \begin{minipage}{0.49\textwidth}    
\begin{algorithm}[H]
   \caption{Generative model over programs}
   \label{programGenerativeModel}
   \begin{algorithmic}
%%      \STATE \textbf{function} sampleProgramFromDSL$(\mathcal{D}, \theta, \tau)$:
%%   \STATE {\bfseries Input:} DSL $\mathcal{D}$, weight vector $\theta$, type $\tau$
%%   \STATE \textbf{Output:} a program whose type unifies with $\tau$
%%   \STATE \textbf{return} sample$(\mathcal{D}, \theta, \varnothing, \tau)$
%% \STATE
     \STATE \textbf{function} sample$(\mathcal{D}, \theta, \mathcal{E}, \tau)$:
  \STATE {\bfseries Input:} DSL $(\mathcal{D},\theta)$, environment $\mathcal{E}$, type $\tau$
  \STATE \textbf{Output:} a program whose type unifies with $\tau$
  \IF{$\tau = \alpha\to\beta$}
  \STATE var $\gets$ an unused variable name
  \STATE body $\sim$ sample$(\mathcal{D},\theta,\{\text{var}:\alpha\}\cup\mathcal{E},\beta)$
   \STATE \textbf{return} \code{(lambda (}var\code{) }body\code{)}
   \ENDIF
   %   \ELSE
   \STATE $\text{primitives} \gets\{p | p: \tau' \in \mathcal{D}\cup\mathcal{E}$ \\
      \hspace*{6.9em}$\text{if }\tau\text{ can unify with yield}(\tau') \} $

   \STATE Draw $e\sim \text{primitives}$, w.p. $\propto\theta_e$ if $e\in \mathcal{D}$ \\
      \hspace*{8.8em}w.p. $\propto\frac{\theta_{var}}{|\text{variables}|}$ if $e\in \mathcal{E}$
   \STATE Unify $\tau$ with yield$(\tau')$.
   \STATE $\left\{\alpha_k \right\}_{k = 1}^K\gets\text{args}(\tau')$ 
%   \STATE unify$(\tau,\beta)$
   \FOR{$k=1$ {\bfseries to} $K$}
 \STATE $a_k\sim\text{sample}(\mathcal{D},\theta,\mathcal{E},\alpha_k)$
 \ENDFOR
 \STATE \textbf{return} \code{(}$e\;a_1\; a_2\; \cdots\; a_K$\code{)}
 \STATE\textbf{where:}
 \STATE yield$(\tau) = \begin{cases}
   \text{yield}(\beta)   &\text{ if }\tau = \alpha\to \beta\\
   \tau   &\text{ otherwise.}
 \end{cases}$ 
 \STATE  args$(\tau) = \begin{cases}
   [\alpha] + \text{args}(\beta)   &\text{ if }\tau = \alpha\to \beta\\
   []   &\text{ otherwise.}
 \end{cases}$
\end{algorithmic}
\end{algorithm}
\vspace{-1.8cm}    \end{minipage}
  \end{wrapfigure}Alg.~\ref{programGenerativeModel} is a procedure for drawing
samples from the generative model $(\mathcal{D},\theta)$.  In practice, we
enumerate programs in order of their probability under  Alg.~\ref{programGenerativeModel} rather than sample them.

@@

We want a recognition model that closely approximates the true posteriors over programs. We formulate this as minimizing the expected KL-divergence, $  \expect\left[\text{KL}\left(\probability[p|x,\mathcal{D},\theta]\|q(p|x) \right) \right]$,
or equivalently maximizing $  \expect\left[\sum_p\probability[p|x,\mathcal{D},\theta]\log q(p|x) \right]$,
%% \begin{equation*}
%%   \expect\left[\sum_p\probability[p|x,\mathcal{D},\theta]\log q(p|x) \right]
%% \end{equation*}
where the expectation is taken over tasks. One could take this expectation
over the observed empirical distribution of tasks,
like how an autoencoder is trained~\cite{hinton2006reducing}; or, one could take this expectation over samples from the generative model, like how a Helmholtz machine is trained~\cite{dayan1995helmholtz}.
We found it useful to maximize both an autoencoder-style objective $\mathcal{L}_{\text{AE}}$ and a Helmholtz-style objective $\mathcal{L}_{\text{HM}}$, giving the  objective for a recognition model, $\mathcal{L}_{\text{RM}} = \mathcal{L}_\text{AE} + \mathcal{L}_\text{HM}$:
\begin{align}\nonumber
\mathcal{L}_{\text{HM}} = \expect_{(p,x)\sim(\mathcal{D},\theta) }\left[\log q(p|x)\right]\quad
\mathcal{L}_{\text{AE}} = \expect_{x\sim X}\left[\sum_{p\in \mathcal{F}_x}
  \frac{\probability\left[x,p|\mathcal{D},\theta \right]}{\sum_{p'\in \mathcal{F}_x}\probability\left[x,p'|\mathcal{D},\theta \right]}\log q(p|x)\right]
\end{align}
The $\mathcal{L}_{\text{HM}}$ objective is essential for data efficiency:
all of our experiments train \system on only a few hundred tasks, which is too little for
a high-capacity neural network $q$.
Once we bootstrap a $(\mathcal{D},\theta)$,
we can draw unlimited samples from $(\mathcal{D},\theta)$
and train $q$ on those samples.

@@

We define a prior distribution over DSLs which penalizes the sizes of the $\lambda$-calculus expressions in the DSL, and put a Dirichlet prior over the weight vector:
\begin{equation}
    \probability[\mathcal{D}]\propto\exp\left(-\lambda\sum_{p\in \mathcal{D}}\text{size}(p) \right)\hspace{1cm}  \probability[\theta|\mathcal{D}] = \text{Dir}(\theta|\alpha)
  \end{equation}
where $\text{size}(p)$  measures the size of the syntax tree of program $p$,
$\lambda$ acts as a regularizer on the size of the DSL,
and $\alpha$ is a concentration parameter controlling the smoothness of the prior over $\theta$.
%Alg.~\ref{grammarInductionAlgorithm} specifies the DSL induction algorithm.

\begin{wrapfigure}{R}{0.6\textwidth}\vspace{-0.7cm}
  \begin{minipage}{0.6\textwidth}    
    \begin{algorithm}[H]
      \caption{The \system Algorithm}
      \label{mainAlgorithm}
      \begin{algorithmic}
        \STATE {\bfseries Input:} Initial DSL $\mathcal{D}$, set of tasks $X$, iterations $I$
        \STATE \textbf{Hyperparameters:} Enumeration timeout $T$
        %     \STATE \textbf{Output:} DSL $\mathcal{D}$, weight vector $\theta$, recognition model $q(\cdot)$
        \STATE Initialize $\theta\gets \text{uniform}$ %, $q_0(\cdot ) = \theta_0$
        \FOR{$i=1$ {\bfseries to} $I$}
        %     \FOR{$x:\tau\in X$}
        \STATE  $\mathcal{F}^{\theta}_x\gets \{p| p\in \text{enum}(\mathcal{D},\theta,T)\text{ if }\probability[x|p] > 0\}$ \footnotesize{(\textbf{Explore})}
        \STATE $q\gets \text{train recognition model, maximizing }\mathcal{L}_{\text{RM}}$ \hspace{0.2cm}(\footnotesize{\textbf{Compile}})
        \STATE  $\mathcal{F}^{q}_x\gets\{p|p\in \text{enum}(\mathcal{D},q(x),T)\text{ if }\probability[x|p] > 0\}$ \footnotesize{\textbf{(Explore)}}

        \STATE $\mathcal{D},\theta\gets $induceDSL$(\{\mathcal{F}^{\theta}_x\cup\mathcal{F}^{q}_x\}_{x\in X})$  \hspace{1.1cm} \footnotesize{\textbf{(Compress)}}
        %% \STATE Define $Q_x(z) \propto \begin{cases}
        %%   \probability[x|z]\probability[z|\mathcal{D}_i,\theta_i]&x\in \mathcal{F}_x\\
        %%   0&x\not \in \mathcal{F}_x
        %% \end{cases}
        \ENDFOR
        \STATE \textbf{return} $\mathcal{D},\theta,q$
      \end{algorithmic}
    \end{algorithm}
  \end{minipage}
\vspace{-0.7cm}\end{wrapfigure}To appropriately score each proposed $\mathcal{D}$ we must reestimate the weight vector $\theta$.
Although this problem may seem 
very similar to estimating the parameters of a probabilistic context free grammar (PCFG),
for which we have effective approaches like the Inside/Outside algorithm~\cite{international2000derivation},
our DSLs are context-sensitive due to the presence of variables
in the programs and also due to the polymorphic typing system.
In the Supplement we derive a tractable MAP estimator for $\theta$.

\begin{figure*}
\centering  \begin{tabular}{ll}
    \toprule
    Example programs in frontiers&Proposed subexpression\\\midrule
    \begin{tabular}{l}
      \code{(lambda (a b) (fold b (cons "," a)}\\
      \phantom{\code{(lambda }}\code{(lambda (x z) (cons x z))))}\\
      \code{(lambda (a b) (fold a b (lambda (x z) (cons x z))))}
    \end{tabular}&
    \begin{tabular}{l}
      \code{(fold a b (lambda (x z) }\\
      \phantom{\code{(fold a b }}\code{(cons x z)))}
      \end{tabular}
\\\bottomrule
  \end{tabular}
  \caption{The DSL induction algorithm proposes subexpressions of programs to add to the DSL.
    These subexpressions are taken from programs in the frontiers (left column), and can introduce new variables (right column: \code{a} and \code{b}). Here, the proposed subexpression appends two lists.} 
  \label{fragmentExample}
\end{figure*}
Alg.~\ref{mainAlgorithm} describes how we combine program search,
recognition model training, and DSL induction.
\begin{comment}
\subsection{Implementing \system}

Alg.~\ref{mainAlgorithm} describes how we combine program search,
recognition model training, and DSL induction.
We note the following implementation details:
(1) We perform an exploration cycle before each compression and compilation  cycle.
(2) On the first iteration, we do \emph{not} train the recognition model on samples from
the generative model because a generative model has not yet been learned --
we instead train the network to only maximize $\mathcal{L}_{\text{AE}}$.
%% (3) Because the frontiers can grow very large,
%% we only keep around the top $10^4$ programs $p$ in each frontier $\mathcal{F}_x$
%% with the highest likelihood $\probability[x,p|\mathcal{D},\theta]$.
(3) During both DSL induction and neural net training, we calculate Eq.~\ref{AIC} and $\mathcal{L}_{\text{RM}}$
by only summing over the top $K$
programs in $\mathcal{F}_x$ as measured by $\probability[x,p|\mathcal{D},\theta]$ -- we found that $K = 2$ sufficed.
(4) For added robustness, we enumerate programs from both the generative model $(\mathcal{D},\theta)$ and the recognition model.
\end{comment}
